---
title: "Multiple Regression on WIC Program"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'D:/R/work directory')
```

```{r working directory, include=FALSE}
setwd("D:/R/work directory")
```

#### import the dataset
```{r pre-processing}
good <- read.csv("good.csv")
good <- na.omit(good[,c("CHID",
                        "mathraw97",
                        "AGE97",
                        "faminc97",
                        "bthwht",
                        "WICpreg",
                        "HOME97")])
attach(good)
```

### Descriptive Statistics
```{r data basics}
summary(good[,c("mathraw97",
                "AGE97",
                "faminc97",
                "bthwht",
                "WICpreg",
                "HOME97")])

sd(mathraw97)
sd(AGE97)
sd(faminc97)
sd(bthwht)
sd(WICpreg)
sd(HOME97)

table(bthwht)
table(WICpreg)

cor(good[,c("AGE97",
            "faminc97",
            "bthwht",
            "WICpreg",
            "HOME97")], use="complete.obs")
```

Build the multiple regression analysis.
```{r model1}
good1 <- good[,c("mathraw97",
                 "AGE97",
                 "faminc97",
                 "bthwht",
                 "WICpreg")]
model1 <- lm(mathraw97 ~ AGE97+faminc97+bthwht+WICpreg, data=good1)
```

### Diagnostics & Corrections
#### Check for Linearity
Check linearity with scatter plot matrix.
```{r}
pairs(good1, panel=panel.smooth)
```

Examine the relationship:
```{r, include=FALSE}
library(ggplot2)
```
1. AGE97 & mathraw97
```{r}
ggplot(good1, aes(x=AGE97, y=mathraw97)) +
  geom_point(size=0.6) +
  xlab("Respondent's Age") +
  ylab("Math Test Score") +
  theme_bw() +
  geom_smooth(method="loess")
```
2. faminc97 & mathraw97
```{r}
ggplot(good1, aes(x=faminc97, y=mathraw97)) +
  geom_point(size=0.6) +
  xlab("Respondent's Family Income") +
  ylab("Math Test Score") +
  theme_bw() +
  geom_smooth(method="loess")
```

#### Check for Homoscedasticity
```{r}
plot(model1)  #enter to get to residual plot 
```
1. model residuals
```{r}
good.res<-resid(model1)
```
2. fitted values
```{r}
fitted.res<-fitted(model1)
plot(fitted.res,good.res)
abline(0, 0, lty=2, col="grey") 
lines(lowess(good.res ~ fitted.res), col="red")
```

#### Check for Normality
1. AGE97
```{r}
#probability plots
hist(AGE97, freq=FALSE)
lines(density(na.omit(AGE97)))
#quantile-quantile plots
qqnorm(AGE97)
qqline(AGE97,col="red")
```
2. faminc97
```{r}
#probability plots
hist(faminc97,freq=FALSE)
lines(density(faminc97))
#quantile-quantile plots
qqnorm(faminc97)
qqline(faminc97,col="red")
```
3. Residuals
```{r}
hist(good.res,10)
boxplot(good.res,main="Boxplot of residuals") 
```

#### Check for Omitted Relevant Variables
*Using Added Variable Plots*
```{r model}
good2 <- good[c("mathraw97",
                "AGE97",
                "faminc97",
                "bthwht",
                "WICpreg",
                "HOME97")]
# mathraw97
model_math <- lm(mathraw97 ~ AGE97+faminc97+bthwht+WICpreg, data=good2)
# HOME97
model_home <- lm(HOME97 ~ AGE97+faminc97+bthwht+WICpreg, data=good2)
```
Compare plotted residuals.
```{r}
plot(density(resid(model_math)))
plot(density(resid(model_home)))
qqnorm(resid(model_math))
qqline(resid(model_math), col="red")
qqnorm(resid(model_home))
qqline(resid(model_home), col="red")

plot(model_home$residuals, model_math$residuals)
abline(lm(model_math$residuals ~ model_home$residuals), col="red")
lines(lowess(model_math$residuals ~ model_home$residuals), col="blue")
```

### Data Transformation
1. Centering AGE97
```{r}
#Create a new centered AGE97 variable.
good2$AGE97c <- good2$AGE97-mean(good2$AGE97)
#Create a new centered AGE97 squared variable
good2$AGE97c2 <- (good2$AGE97c**2)
```
2. Log-Transform faminc97
```{r}
good2$logfaminc <- ifelse(good2$faminc97 <= 1, 0, 
                          ifelse(good2$faminc97 > 1, log(good2$faminc97), NA))
```

Create a new model after transformation.
```{r model2}
model2 <- lm(mathraw97 ~ AGE97c+AGE97c2+logfaminc+bthwht+WICpreg+HOME97, data=good2)
```

Extract the model fitted values and residuals.
```{r}
pred1 <- as.data.frame(model2$fitted.values)
pred1$residuals <- (model2$residuals)
```

Produce a scatter plot of residuals against IVs.
```{r}
plot(good2$logfaminc, pred1$residuals)
abline(lm(pred1$residuals ~ good2$logfaminc), col="red")  
lines(lowess(good2$logfaminc, pred1$residuals), col="blue")

plot(good2$HOME97, pred1$residuals)
abline(lm(pred1$residuals~good2$HOME97), col="red")  
lines(lowess(good2$HOME97, pred1$residuals), col="blue")

summary(model2)
```

### Outlier Examination
```{r model3}
good3 <- good

good3$AGE97c <- good3$AGE97-mean(good3$AGE97)
good3$AGE97c2 <- (good3$AGE97c**2)
good3$logfaminc <- ifelse(good3$faminc97 <= 1, 0, 
                          ifelse(good3$faminc97 > 1, log(good3$faminc97), NA))

model3 <- lm(mathraw97 ~ AGE97c+AGE97c2+logfaminc+bthwht+WICpreg+HOME97, data=good3)
```

Create a new data frame for outlier information.
```{r}
outliers <- good3[,c("CHID",
                     "mathraw97",
                     "AGE97c",
                     "AGE97c2",
                     "logfaminc",
                     "bthwht",
                     "WICpreg",
                     "HOME97")]
outliers$r <- rstudent(model3)
outliers$lev <- hatvalues(model3)
outliers$cd <- cooks.distance(model3)
outliers$dffit <- dffits(model3)
```

Create a separate data frame with the dfbetas.
```{r}
dfbetaR <- dfbetas(model3)
```
Check the order of the variable names.
```{r}
colnames(dfbetaR)
```
Assign new variable names.
```{r}
colnames(dfbetaR) <- c("int_dfb","agec_dfb","agec2_dfb",
                       "income_dfb","bthwht_dfb",
                       "wic_dfb","home_dfb")
outliers<-cbind(outliers,dfbetaR)
head(outliers)
```

1. Discrepency
```{r}
#abs(standardized resid) > 2 (or 3)
plot(outliers$r, 
     xlab="Index",
     ylab="studentized residuals",
     pch=19)
#Add fit lines.
abline(lm(r~CHID, data=outliers), col="red")
abline(h=2, col="blue")
abline(h=-2, col="blue")
abline(h=3, col="green")
abline(h=-3, col="green")

rstudent1 <- subset(outliers,abs(r)>3)
View(rstudent1)
```

2. Leverage
```{r}
#leverage > (2k+2)/N 
plot(outliers$lev, 
     xlab="Index",
     ylab="leverage",
     pch=19)
#Add fit lines.
abline(lm(lev~CHID, data=outliers), col="red")
abline(h=0.007, col="blue")
abline(h=0.01, col="green")

leverage1 <- subset(outliers,lev> .01)
View(leverage1)
```

3. Influence (Cook's D)
*Cook's D > 4/N* 
```{r model4}
model4 <- lm(mathraw97 ~ AGE97c+AGE97c2+logfaminc+bthwht+WICpreg+HOME97, data=subset(outliers, cd<(4/2042)))
summary(model4)
```

Maybe our critical cooks.d value of 4/2042 is too strict.
```{r}
large_cd <- subset(outliers, cd >(4/2042))
```

Next, produce a histogram and print the sample quantiles of our Cook's D outlier measure to the console
```{r}
library(Hmisc)
describe(large_cd$cd)
hist(large_cd$cd)
quantile(large_cd$cd, probs = seq(0, 1, 0.05))
```

Take a closer look at these observations
```{r}
large_cd2<-subset(outliers, cd > 0.007008524)
#examine observations in this object
View(large_cd2)
```

### Adjusted Model
```{r model5}
outliers2<-subset(outliers, cd < 0.007008524)

#Now you can compare the results produced from our model1 model with the results
#produced with our new dataset.
model5 <- lm(mathraw97 ~ AGE97c+AGE97c2+logfaminc+bthwht+WICpreg+HOME97, data=outliers2)

summary(model5)

detach(good)
```
